Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-08-25 Manuscript/domReview.qmd
Big changes from ESA travel period
Added approx 30 articles, re-wrote manuscript draft

abstract: | 
  Dynamic occupancy models are an increasingly important tool for the analysis of wildlife occurence data due to their favourable balance of complexity and ease of implementation. Uses of this model class are diverse, including explorations of the drivers of occupancy as well as predictions of species distributions across landscapes and into the future. The application of these models does require careful navigation of the modelling process including important decisions on structure, covariate inclusion, and assessment of model performance.

  In this article we present a systematic review of implementations of the dynamic occupancy model, with particular emphasis on how authors have approached the modelling process. Our findings indicate that authors have found this model class suitable for a considerably diverse set of ecological questions and have successfully fit models to data collected under a wide range of conditions, highlighting the high degree of flexibility available. However, the manner of application appears to be quite variable: approaches to covariate selection vary considerably, and model complexity generally speaking is low. The implications of these decisions as they relate to model outputs are yet to be explored in detail and should be a priority for future research.

embed-resources: true
bibliography: references.bib
---

@@

Since its introduction two decades ago by @mackenzie2003 the dynamic multi-season occupancy model has become an important tool for the analysis of species occurrence throughout space and time. In the ecological modelling landscape dynamic occupancy models (DOMs) occupy a favourable position between model complexity and ease of data collection. In explicitly estimating the processes which guide occupancy while accounting for imperfect detection, the model class offers more advanced insights than simpler static species distribution models; despite this, input data requirements remain achievable compared to other more complex process-explicit models.

The use-cases of this model class are wide-ranging, and overlap considerably with other models such as species distribution models [@elith2009] and stochastic patch occupancy models [@gutiérrez-arellano]. Estimates of occupancy through time are independently valuable for many monitoring programs, and DOMs are a particularly favourable choice when species are not detected perfectly [@guillera-arroita2017]. DOMs may also be used in testing specific hypotheses of relationships between occupancy and the environment, or in identifying the drivers of occupancy in a system. Although infrequently applied thus far, a promising application of DOMs is their use to make predictions to unsurveyed sites or into the future [@briscoe2021; @kéry2013a]. In each of these use cases, authors must navigate the modelling process and make important decisions on aspects of model structure, the process of determining which covariates to incorporate, and the critical assessment of model outputs such as parameter estimates or predictions.

This review seeks to provide an overview dynamic occupancy model's first two decades of use. We discuss the DOMs key features, limitations, and implementation process, before presenting a systematic review of articles implementing the model. Based on these results, we present key considerations for authors seeking to build these models in their work and outline priorities and opportunities for future research involving this flexible and important model class.

## Model form

Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied. The basic model form as described by @mackenzie2003 is depicted in @fig-modelform. Under this parameterisation, sites exist in either occupied or unoccupied states and may transition between these states between each time-step. Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction. Given that a site is occupied, the probability of observing the species during a survey is given by the detection parameter. Under this formulation, it is assumed that false-positive detections do not occur.

![The original parameterisation of the dynamic occupancy model as described by @mackenzie2003. The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}

To disentangle the ecological and observation processes and estimate the probability of detection, the DOM relies on a hierarchical sampling design based on Pollock's robust design [@pollock1982] and depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy. Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy. Most frequently these observations are collected by revisiting the site on separate occasions, although alternative methods to generate repeat observations exist [@mackenzie2005]. These alternatives include surveys conducted at multiple locations within a site, use of multiple observers during a survey, or recording of the time elapsed before a detection is recorded.

![The sampling design of the dynamic occupancy model.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}

## Key assumptions

Dynamic occupancy models require a small number of assumptions with important implications for users.

1.  **False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species. Several authors have commented on the perils of not accounting for false positives where they occur [@miller2015; @mcclintock2010]. Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the intended definition of occupancy - in the authors' case, transient individuals where the authors were interested in breeding occupancy. Extensions to the dynamic occupancy model which account for false positive error are available [@royle2006; @miller2011; @miller2013] and should be carefully considered where this assumption cannot be met.

2.  **Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and departure periods between sites. More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.

3.  **No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of modelled heterogeneity on parameter estimates [@mackenzie2017]. Environmental factors are near-guaranteed to influence species occupancy and observation in any system, particularly at larger spatial scales. While studies should attempt to incorporate the key drivers of occupancy via inclusion of covariates, there is limited discussion on the best methods to achieve this and the implications of failure to capture this variation.

## Model extensions

The DOM has been frequently extended over the years, with alternate parameterisations available to better address many needs. In addition to the false-positive implementations previously discussed, prominent model extensions include multi-state DOMs which allow sites to exist in states other than occupied and unoccupied [@nichols2007], multi-scale models describing occupancy at multiple spatial scales [@nichols2008], and multi-species models permitting joint estimation of species with interactions [@devarajan2020]. For an in-depth discussion of additional model extensions and their applications see @bailey2014.

# Methods

## Article collection

This systematic review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003. To qualify for inclusion articles must have fit a model which fulfils these criteria: i) describes multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) has multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; and iv) uses real (i.e., not simulated) field ecological data.

Papers were gather for inclusion in two stages.

1.  In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were entered into Google Scholar. Up to 100 articles were downloaded for each string in order of appearance. Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021. For each combination of search string and year strata, a random 25% or 5 articles (whichever was greater) was processed. Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is best interpreted as a distinct model class**.**

2.  To increase sample size a secondary sample of articles was generated, using the complete list of articles which have cited @mackenzie2003 (downloaded via Web of Science). Articles in this list were split into the same year strata, plus an additional 2022-2023 category. 15 articles were drawn from each strata and again processed with replacement from their strata.

## Article processing

For each included article data was collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation. A full spreadsheet of data collected is given in **Appendix I.**

### Authorship

To assess patterns of authorship, the location of each articles first author's first-listed affiliation was denoted. For *all* authors[,]{.underline} each affiliation was denoted as one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.

### Objectives

For each article five non-exclusive research objectives were indicated based on the authors' stated aims: 'Estimating parameters,' where authors express interest in estimates of site occupancy, colonisation, or extinction; 'Testing hypotheses,' where authors explore *specific* relationships between covariates and model parameters; 'identifying drivers,' where authors attempt to find *which* covariates influence model parameters; 'generating predictions,' where which authors estimate model parameters either for unsurveyed sites or under possible future conditions; and 'methods development,' where authors introduce or test aspects of dynamic occupancy models.

### Taxa and study areas

Focal taxa were divided into broad categories (bird, mammal, etc.). Taxa were denoted as threatened either where the authors refer to them as so or where the species is listed on the IUCN red list (IUCN 2024) - this determination was made to account for unlisted sub-species or species of local concern. Study areas are defined as the intended area of inference for a given study in which all sites are contained; size of these areas are reported as orders of magnitude in square kilometres to partly account for uncertainty in reporting.

### Covariates

Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models. For each parameter all covariates were recorded in broad mutually-excluded categories. For each of these covariates, data was collected on whether the covariate was remotely collected or required direct observation, whether polynomial terms were considered, whether interaction terms included this covariate were considered, and whether the covariate was dynamic – i.e, varied between seasons.

### Model selection and evaluation

To describe the complex model selection process, non-exclusive categories were applied to each article including: 'A priori,' where only one model is considered; 'candidate suite,' where a predefined set of models are considered; 'procedural,' where covariates are selected parameter-by-parameter; 'exhaustive,' where all possible model combinations are fit; 'simple precursors,' where selection is predicated on results of another, simpler model implementation; and 'other.'

Model evaluation was similarly categorised - options in this case included 'relative fit,' for relative metrics of fit such as AIC; 'model averaging,' where model weights are applied to candidate models; 'goodness of fit,' where absolute metrics of fit are used; 'out-of-sample validation' and 'in-sample-validation,' where predictive performance is evaluated using in-sample or out-of-sample data respectively.

# Results

A total of 84 articles were included in this review, 47 from the preliminary sample and 37from the secondary sample. Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles citing @mackenzie2003 would have met inclusion criteria.

![Estimated coverage of articles included in this review, estimated based upon the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles were published with first-authors from a total of 15countries, with authors from the United States producing 64% of articles. 86% of articles had at least one author based at an academic institution, 46% had authors at government institutions, and 25% had authors from non-profit organisations.

## Study locations, focal species, and data collection

![An overview of study areas, focal species, and data collection for reviewed articles.](Figures/ReviewBox.png){#fig-StudyDetails}

Data for DOMs were collected from study areas on all continents and in 21 countries, with studies in the United States again representing a majority of the total articles published. These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).

Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates. A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017. 35% of articles modelled threatened taxa, and 10% modelled species invasive to the study area. The majority of articles (50, 60%) fit single-species DOMs to one taxa. Of the remainder, 26 (31%) fit independent single-species models to more than one taxa. 8 articles fit (10%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).

Data used to fit dynamic occupancy models included human survey data such as transects or point counts (69% of articles), live-trapping data (20%), camera traps (12%) or bioacoustic monitors (1%), and citizen science data (12%); some articles incorporate multiple detection methods to build their detection datasets. Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for 39 months. The number of seasons ranged from 2 to 204 with a median of 5; the number of observations per season ranged from 2 to 153 with a median of 5. The number of sites surveyed per article ranged from 7 to 9394, with the median study including 98 sites (@fig-StudyDetails c).

## Covariates

A summary of covariates considered in the review sample is presented in @tbl-covariates. For each category of covariate relative frequency and characteristics are given. For all categories, the rate of consideration of polynomial terms does not exceed 16%, and the rate of consideration of interactions does not exceed 34%.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "Attributes of covariates found in dynamic occupancy models"

>>>>

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research. Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play. The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates. In the two decades since their introduction the DOM has become a popular tool for a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.

  In this review we examine how ecologists have applied the DOM in its first two decades of use. Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data was collected for them, and how authors have built their models to achieve their objectives. Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades. The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.

bibliography: references.bib
csl: apa.csl
embed-resources: true
---

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy a difficult quantity to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data. Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 offer a powerful framework for modelling presence/absence data while accounting for each of these challenges. Originating in the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly incorporating the key processes which drive occupancy while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception. Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied the DOM to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models. Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.

### Dynamic occupancy model form and assumptions

The DOM unites two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform). Under this parameterisation, sites exist in either an occupied or unoccupied state. In the first time-step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

This description of the DOM model form and sampling design are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003. Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

DOMs make a number of assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.

# Systematic review methods

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles which fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.

For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach. To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims. The possible categories were *Observing trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.

We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable. Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened. This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern. Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all site — this was measured to the order of magnitude to account for uncertainty in reporting.

We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation. To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation. Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data. For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**

# Applications of dynamic occupancy models

A total of 60 articles were included for this review. Based on the acceptance rates within each strata, an estimated 516 of the 1300 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

![Estimated coverage of the articles included in this review. A total of 60 articles from 2000-2023 were processed; bars indicating the estimated number of qualifying articles remaining are based on the rejection rate of each strata and the number of remaining unread articles in the pool. The proportion of articles included from each strata were as follows: 2004-2007, 12.5%; 2008-2011, 21.3%; 2012-2015, 44.4%; 2016-2019, 32.4%; 2020-2023, 57.1%.](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranging from under one month to forty years, with a median of 7 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

DOMs have been used to study a variety of different species, though the majority of studies have been conducted on data for vertebrate taxa. They have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and DOMs have been touted as a valuable tool for modelling disease dynamics. @mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of applications have been fit for terrestrial species, though there have been a limited number of studies on aquatic systems, including @fisher2014's application to invasive salmon, @falke2012's on Great Plains stream fishes, and @pendleton2022's on whale occupancy dynamics.

While most articles in our sample fit a model to a single species, 26 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models used various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicitly interactive models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxa is likely to be reduced for practicality's sake.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 38 articles fit models to only a single taxa, 20 fit independent models to multiple taxa, and 6 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 10 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods. In our sample 76% of articles used human observational data, 19% used live-trapping methods, and 5% used detections from camera trapping or acoustic monitoring. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger sign. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in our only application of a DOM for a marine species @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' treating poachers as species which uses ranger reports to create detection histories.

Notably, many of these studies use data which was not originally collected in a robust design framework for occupancy modelling. In these articles, authors manipulated their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care in estimating detection probability, which is generally expected to vary based on observation method; however, this can be reasonably achieved by including survey method as a covariate on detection probability [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019]. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.

# Practices in implementation and model building

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data is collected, and the ways in which those covariates are statistically represented in DOMs. We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environmental itself. Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of different covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry (31% of articles) and connectivity (26% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails. Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy [@mangan2019] and models with bullfrog covariates on native amphibian occupancy [@rowe2019], each of which effectively incorporate species interactions without fitting explicitly multi-species models.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. We calculate the percentage of articles which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;). We also present the percentage of all implementations of each covariate category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate as a non-linear relationship or as part of an interaction with another covariate."
